---
title: "COVID-19 Data Visualization"
output: 
  learnr::tutorial
runtime: shiny_prerendered
date: '15 March 2020'
author: Nicolò Caporale
---

## Intro

This is a didactic experiment on basic data analysis with R, using the datasets downloaded from the European Centre for Disease Prevention and Control webpage:
https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide

***

This is not a way to underestimate the suffering of this period.

Unfortunately every number here is a person that got sick or died, a tragedy.
So I want to underline that this is not playing with numbers and colors but an attempt to help people to critically look at real data with a scientific perspective.

***

Outside the emergency, our main activity is to study the physiology and pathology of neurodevelopment, in the lab of Giuseppe Testa at the European Institute of Oncology and the Human Technopole, University of Milan, Italy. 
You can read more here http://www.testalab.eu/bibliography/

### SetUp
In this first chunck we start loading the libraries that contain the functions that will facilitate our work later
```{r, message=FALSE}
library(Hmisc)
library(tidyverse)
library(readxl)
library(learnr)
library(RColorBrewer)
data <-  read_excel("DataECDC/COVID-19-geographic-disbtribution-worldwide-2020-03-15.xls")
```

```{r setup, include=FALSE}
library(Hmisc)
library(tidyverse)
library(readxl)
library(learnr)
library(RColorBrewer)
data <-  read_excel("DataECDC/COVID-19-geographic-disbtribution-worldwide-2020-03-15.xls")
col <- topo.colors(length(unique(data$CountryExp)))
names(col) <- unique(data$CountryExp)
data$Date <- as.character(data$DateRep)
```


## Data Check {data-progressive=TRUE}

Here we have a look at the table and the kind of data in front of us.
Understanding the data in front of us is the first necessary step for every kind of data analysis.


### visualize your data

```{r}
DT::datatable(data)
```

### check your data with R

```{r}
dim(data)
colnames(data)
head(rownames(data))
head(data$CountryExp)
head(data$NewConfCases)
```

### summary statistics
This command makes the summary statistics for the object of your interest

```{r}
summary(data)
```

### Exercise {data-allow-skip=TRUE}

*Here's a simple exercise with an empty code chunk provided for entering the answer and make your own personalised analysis*

Write the R code required to view the names of the column of your dataset:

```{r ex1, exercise=TRUE}

```

```{r ex1-solution}
colnames(data)
```

Write the R code required to check which type of data you have in the columns "CountryExp" and "NewConfCases" of the dataset:

```{r ex2, exercise=TRUE}

```


```{r ex2-solution}
class(data$CountryExp)
class(data$NewConfCases)
```


Write the R code required to check summary statistics for the first column of your dataset:

```{r ex3, exercise=TRUE}

```

```{r ex3-solution}
summary(data$DateRep)
```


## Numeric Variable Exploration {data-progressive=TRUE}

### Scatterplot 

The first kind of useful plot for looking at the data is a scatterplot between 2 numeric variables 

```{r}
ggplot(data, aes(x=NewConfCases, y=NewDeaths)) + geom_point()
```

And you can use the boxplot function that shows you the distribution of your data points. By default it highlights also the outliers.

Here you can read more: 
https://towardsdatascience.com/understanding-boxplots-5e2df7bcbd51

```{r}
boxplot(data$NewConfCases)
```


Often when looking at distribution of points, transforming the data with the log function is useful, and you can also color the points for different countries to have more information

```{r, fig.width=8, fig.height=6, warning=FALSE}
ggplot(data, aes(log(NewConfCases+0.1), NewDeaths,)) + 
  geom_point()
```


```{r, warning=FALSE}
boxplot(log(data$NewConfCases+0.1))
```


### Exercise {data-allow-skip=TRUE}

Plot distribution of new deaths with boxplot:

```{r ex4, exercise=TRUE}

```

```{r ex4-solution,warning=FALSE}
boxplot(data$NewDeaths)
boxplot(log(data$NewDeaths+0.1))
```


## Numeric Variable between groups {data-progressive=TRUE}

Those were didactic examples that however do not give us a real understanding of the data.
Now we try to have a closer look at interesting things.

### Histograms
One thing we can explore is the distribution of cases across different groups.
Here for example we use histograms to compare today distribution of new cases across countries

```{r, fig.height=8, fig.width=12}
length(unique(data$CountryExp))
col <- topo.colors(length(unique(data$CountryExp)))
names(col) <- unique(data$CountryExp)

data$Date <- as.character(data$DateRep)

March14 <- data[which(data$Date=="2020-03-14"),]
ggplot(March14, aes(CountryExp, NewConfCases, col= col[March14$CountryExp],)) +
  geom_col()+
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position="none")
```


### Exercise {data-allow-skip=TRUE}

Analyse in a similar way the number of new deaths

```{r ex7, exercise=TRUE, fig.height=8, fig.width=12}

```

```{r ex7-solution, fig.height=8, fig.width=12}
March14 <- data[which(data$Date=="2020-03-14"),]
ggplot(March14, aes(CountryExp, NewDeaths, col= col[March14$CountryExp],)) +
  geom_col()+
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position="none")
```


### 

Another way to perform a similar kind of analysis is to compare distribution of new cases along time, splitting the bars for countries

```{r, fig.height=10, fig.width=10}
ggplot(data, aes(Date, NewConfCases, fill=CountryExp)) +
  geom_col()+
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position="bottom")
  
```

But if there are too many countries this representation is difficult to read, so we move to the next chapter:

## Cases along time {data-progressive=TRUE}

One basic analysis we can do is to check the evolution of new cases along time with scatterplots.

```{r}
ggplot(data, aes(x= DateRep, y= NewConfCases)) + 
  geom_point()+
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))
```


But with all the data aggregated is kind of useless, so we can split and observe the evolution in different Countries
 
```{r, fig.height=6, fig.width=8}
Italy <- data[which(data$CountryExp=="Italy"),]
  p <- ggplot(Italy, aes(x= DateRep, y= NewConfCases, col= col["Italy"])) + 
    geom_point() + 
    ggtitle("Italy") +
    theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
    theme(legend.position = "none") 
  p

China <- data[which(data$CountryExp=="China"),]
  p <- ggplot(China, aes(x= DateRep, y= NewConfCases, col= col["China"])) + 
    geom_point() + 
    ggtitle("China") +
    theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
    theme(legend.position = "none") 
  p
  
Germany <- data[which(data$CountryExp=="Germany"),]
  p <- ggplot(Germany, aes(x= DateRep, y= NewConfCases, col= col["Germany"])) + 
    geom_point() + 
    ggtitle("Germany") +
    theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
    theme(legend.position = "none") 
  p
```

###

And now we will find a way to do the same for all the country in a fast way:

```{r, fig.height=6, fig.width=8}
for (i in unique(data$CountryExp)) {
  tmp <- data[which(data$CountryExp==i),]
  p <- ggplot(tmp, aes(x= DateRep, y= NewConfCases, col= col[i])) + 
    geom_point() + 
    ggtitle(i) +
    theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
    theme(legend.position = "none")  
  print(p)
}
```

### Exercise {data-allow-skip=TRUE}

Repeat the analysis for the number of deaths across EU and non-EU territories. 

```{r ex5, exercise=TRUE}

```

```{r ex5-solution}
for (i in unique(data$EU)) {
  tmp <- data[which(data$EU==i),]
  p <- ggplot(tmp, aes(x= DateRep, y= NewDeaths)) + 
    geom_point() + 
    ggtitle(i) +
    theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
    theme(legend.position = "none") 
  print(p)
}
```


## Regression {data-progressive=TRUE}

Now we perform a linear regression to follow the trend

```{r, warning=FALSE, fig.width=8, fig.height=7}
Italy <- data[which(data$CountryExp=="Italy"),]

Italy$pred <- predict(lm(NewConfCases ~ DateRep , data = Italy))
ggplot(Italy, aes(x =DateRep, y = NewConfCases)) +
  geom_point() + 
  geom_line(aes(y = pred)) +
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position = "none") 
```

Or better fit an exponential model in this case

```{r, warning=FALSE,fig.width=8, fig.height=7} 
Italy$pred <- exp(predict(lm(log2(NewConfCases+0.01) ~ DateRep , data = Italy)))
ggplot(Italy, aes(x =DateRep, y = NewConfCases)) +
  geom_point() + 
  geom_line(aes(y = pred)) +
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position = "none") 
```

### Exercise {data-allow-skip=TRUE}

Now repeat the same analysis for the cases of new deaths along time for Germany

```{r ex6, exercise=TRUE,fig.width=8, fig.height=7}

```

```{r ex6-solution,fig.width=8, fig.height=7}
Germany <- data[which(data$CountryExp=="Germany"),]
Germany$pred <- exp(predict(lm(log2(NewDeaths+0.01) ~ DateRep , data = Germany)))
ggplot(Germany, aes(x =DateRep, y = NewDeaths)) +
  geom_point() + 
  geom_line(aes(y = pred)) +
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position = "none")  
```


You can read more here
http://r-statistics.co/Linear-Regression.html



## Total Cases {data-progressive=TRUE}

Finally we compute the total number of cases for each country and look again at the data

```{r, fig.width=12, fig.height=8}
data$totalCases <- 0
data$totalDeaths <- 0

for (i in unique(data$CountryExp)) {
data$totalCases[which(data$CountryExp==i)] <- sum(data$NewConfCases[which(data$CountryExp==i)])  
}


ggplot(data, aes(CountryExp, totalCases, col=col[data$CountryExp])) +
  geom_col()+
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position="none")

ggplot(data, aes(CountryExp, log(totalCases+0.1), col=col[data$CountryExp])) +
  geom_col()+
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position="none")

```


### Exercise {data-allow-skip=TRUE}

Analyse in a similar way the number of deaths

```{r ex8, exercise=TRUE,fig.width=12, fig.height=8}

```

```{r ex8-solution, fig.width=12, fig.height=8}
for (i in unique(data$CountryExp)) {
data$totalDeaths[which(data$CountryExp==i)] <- sum(data$NewDeaths[which(data$CountryExp==i)])  
}

ggplot(data, aes(CountryExp, totalDeaths, col=col[data$CountryExp])) +
  geom_col()+
  theme(axis.text.x= element_text(angle = 90, hjust = 1, vjust = 1))+
  theme(legend.position="none")
```



```{r child='footer.Rmd'}
```
